An Object-Logic Explanation for the Inconsistency in G\"odel's Ontological Theory

### 2016-07-10 Papers/2016/KI/main.tex
Improve section about explanation

\section{Intuitive Inconsistency Argument} \label{sec:inconsistency}

@@

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property $\lambda x. \bot$ (also denoted $\emptyset$, as in HOL it is customary to think of unary predicates as sets).
%
Note that the terms for the empty property\footnote{An additional lambda abstraction occurs in the empty property in \textsc{Leo-II}'s proof (and also in the reconstruction in Isabelle) because the embedding approach lifts the boolean type $o$ to $\iota \imp o$.} ($\lambda x. \bot$) and for the property of self-difference ($\lambda x.  x\not=x$) have identical denotations in a logic setting
with functional and Boolean extensionality as assumed
here. Nevertheless, some philosophers\footnote{Private communication with Andr\'e Fuhrmann.} may actually prefer the use of
self-difference over the empty property in
the analysis below. However, for the proof to go through it is
irrelevant which notion we use and the reader may simply replace the
empty property by self-difference.

% Philosophers
% might actually prefer using the latter variant over the
% former. However, on the given logic setting, with functional and
% Boolean extensionality, both terms have identical
% denotations. Consequently, they are mutually replacable below and the
% reader may simply use the version he prefers.

\paragraph{Informal Argument.} \label{sec:arg1}
Based on the hints found in \textsc{Leo-II}'s refutation, we conceived the following informal explanation for the inconsistency of G\"odel's axioms:

@@

\item To see that the sentence above is contradictory, we may reason semantically, thinking of possible worlds. If $w_0$ is the arbitrary current world, the $\pos$ operator forces the existence of a world $w$ accessible from $w_0$ such that $\nec \bot$ is true in $w$. But $\nec \bot$ can only be true in $w$, if there is no world $w'$ accessible from $w$. In logics with a reflexive or symmetric accessibility relation (e.g. \KB), it is easy to see that there must be a world $w'$ accessible from $w$: either $w'$ itself, in case of a reflexive relation, or $w_0$, in case of a symmetric relation. In fact, even in \K, with no accessibility condition, there must be a world $w'$ accessible from $w$. The reason is that $\pos \nec \bot$ should be \emph{valid} (true in all worlds). Therefore, it is true in $w$ as well, where the existence of an accessible world $w'$ is forced by the $\pos$ operator. As a model for $\pos \nec \bot$ (which is a consequence of G\"odel's axioms) cannot be built, G\"odel's axioms are inconsistent.
\end{enumerate}

>>>>

\section{Intuitive Explanations for the Inconsistency} 
\label{sec:inconsistency}

@@

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property\footnote{
  Note that the terms for the empty property ($\lambda x. \bot$) and for the property of self-difference ($\lambda x.  x\not=x$) have identical denotations in the logic setting
  with functional and Boolean extensionality assumed
  here. 
  %Nevertheless, some philosophers may actually prefer the use of
  %self-difference over the empty property in
  %the analysis below. 
  For the proof to go through it is
  irrelevant which property is used.
} $\lambda x. \bot$ (also denoted $\emptyset$, as in HOL it is customary to think of unary predicates as sets)\footnote{An additional lambda abstraction occurs in the empty property in \textsc{Leo-II}'s proof (and also in the reconstruction in Isabelle) because the embedding approach lifts the boolean type $o$ to $\iota \imp o$.}. Based on this hint, we conceived the following informal explanation for the inconsistency of G\"odel's axioms (reproduced without change from \cite{C55}):
%

@@

\item To see that the sentence above is contradictory, we may reason semantically, thinking of possible worlds. If $w_0$ is the arbitrary current world, the $\pos$ operator forces the existence of a world $w$ accessible from $w_0$ such that $\nec \bot$ is true in $w$. But $\nec \bot$ can only be true in $w$, if there is no world $w'$ accessible from $w$. In logics\footnote{
  Interestingly, the refutation automatically generated by
  \textsc{Leo-II} uses a symmetric accessibility relation, and thus
  requires the modal logic \KB. The informal, human-constructed
  refutations described here, on the other hand, requires only the
  weaker modal logic \K. In our experiments \textsc{Leo-II} (like all
  other HOL provers) was still too weak to automatically prove the
  inconsistency already in logic \K. Hence, this remains an open problem for automated
  theorem provers.
} with a reflexive or symmetric accessibility relation (e.g. \KB), it is easy to see that there must be a world $w'$ accessible from $w$: either $w'$ itself, in case of a reflexive relation, or $w_0$, in case of a symmetric relation. In fact, even in \K, with no accessibility condition, there must be a world $w'$ accessible from $w$. The reason is that $\pos \nec \bot$ should be \emph{valid} (true in all worlds). Therefore, it is true in $w$ as well, where the existence of an accessible world $w'$ is forced by the $\pos$ operator. As a model for $\pos \nec \bot$ (which is a consequence of G\"odel's axioms) cannot be built, G\"odel's axioms are inconsistent.
\end{enumerate}
